Solution of hydrodynamic equilibrium in black-oil reservoirs: the classical material balance equation

### 2015-01-23 pituba/article/article_spe.tex
news

The equation developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, phase or thermal equilibrium. An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}. The novelty on hydrodynamics was the introduction of a macroscopic equation of motion using Darcy's law.

@@

 The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here. This approach extends the material balance equation to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, requires negligible non-equilibrium effects in phase equilibrium.

@@

The material balance equation is widely used to relate production and injection to reservoir pressure. Actually, all averaged phase properties in Eq.~\eqref{eq: MBE} depend on the pressure field, and so does the pore volume. If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.

@@

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume hydrostatic equilibrium: $\vec{u}_\alpha \equiv 0$. This implies that $\nabla p_\alpha \equiv \rho_\alpha \vec{g}$, and, as a consequence, pressure becomes a function of time and depth given by the following expression:
\begin{equation}\label{eq: p=rhogh}
\int_{p_\alpha (t,z_{datum})}^{p_\alpha (t,z)} \frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - z_{datum}\right) \, ,
\end{equation}
where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha (t,z_{datum})$ is the pressure at the datum level $z_{datum}$.
It is now possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.

It can be shown that neglecting the variation of density with depth does not introduce  important errors. As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. For instance, the average oil density can be approximated to the local density at the datum level: $\bar{\rho}_o (t) \approx \rho_o (t, z_{datum})$.

@@

This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase,
\begin{equation}\label{eq: grado}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o \, ,
\end{equation}
 and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}\label{eq: gradodz}
 \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

@@

Assuming reservoir temperature to be sufficiently above gas critical temperature and thus a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

@@

Oil and gas densities are thus approximately constant with depth.
The datum level for computing phase density (or the formation volume factor) can be any arbitrary depth inside phase volume. The pressure calculated in the material balance equation corresponds to the local pressure at $z_{datum}$ for negligible hydrodynamic gradients.

@@

For the case of an oil reservoir with a gas cap, the datum level can be defined at the gas-oil contact. The point here is that at gas-oil contact both gas and oil densities are computed using the pressure local value. If the datum level, common to all phases, is otherwise defined below gas-oil contact, a higher gas density will be computed because of the oil hydrostatic gradient:
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the gas hydrostatic gradient used in Eq.~\eqref{eq: gradgdz}.

It was shown that the pressure variation with depth is not important for the computation of gas, oil and water densities. Nevertheless, pressure also plays a second role in the material balance equation by defining phase equilibrium, for what the pressure variation with depth may not be negligible in the solubility ratio: 
\begin{equation}\label{eq: dRsdz1}
\frac{1}{R_s}\frac{d R_s}{dz} = \frac{1}{R_s}\frac{d R_s}{dp_o} \frac{dp_o}{dz} = \frac{\rho_o g}{R_s}\frac{dR_s}{dp_o} \, .
\end{equation}

@@

\section{Solution of material balance equation}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We shall start defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most. Equivalently, we can say that only one solution exists for the material balance equation.

@@

The residual function clearly satisfies the first two affirmatives because fluid properties are continuous everywhere and are also differentiable everywhere except for the first order phase change at bubble-point pressure.
The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the solid rock). For the fluid phases, each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

\textit{In-situ} volumes (e.g. $N-N_p$ for oil phase) are non-negative values. Pressure derivatives of fluid volumes are non-positive values (fluid volume cannot increase as pressure increases) and pressure derivative of porous volume is a positive value (as fluid pressure increases, porous volume expands owing to rock compression).

The value of $d\varepsilon/dp$ is thus negative if at least one phase is compressible because it would be a sum of products between negative and positive factors.

@@

It must be observed that the result of gas reentering oil phase completely neglects non-equilibrium effects, e.g. compositional gradients, and must be used with care. For instance, it may be in strong disagreement with real reservoir behavior when gravity segregation is present.

>>>>

The equation developed by \cite{schilthuis1936active} did not account for the non-existence of mechanical, chemical or thermal equilibrium. An attempt to introduce non-equilibrium effects of hydrodynamic origin came with \textit{black-oil} equations \citep{aziz1979petroleum,blackoil}. The novelty on hydrodynamics was the introduction of a macroscopic equation of motion using Darcy's law.

@@

 The same approach of varying bubble-point pressure widely used in reservoir simulation, which is found in \cite{aziz1979petroleum} and \cite{ertekin2001basic}, is also applied here. This approach extends the material balance equation to saturated reservoirs under pressure maintenance (e.g. fluid injection). The applicability, however, requires negligible non-equilibrium effects in chemical equilibrium.

@@

The material balance equation is widely used to relate production and injection to reservoir pressure. Actually, all averaged phase properties in Eq.~\eqref{eq: MBE} depend on the pressure field. If the reservoir is under production, though, the pressure field is not known and for its prediction one would have to resort to more advanced techniques such as reservoir simulation.

@@

To overcome this difficulty, a possible simplification to the pressure field is to neglect hydrodynamic gradients and assume mechanical equilibrium, i.e., $\vec{u}_\alpha = 0$. This implies that pressure at some time instant is only a function of depth: $\nabla p = \partial p_\alpha/\partial z = \rho_\alpha \vec{g}$.

% so that pressure is a function of time and depth given by the following expression:
%\begin{equation}\label{eq: p=rhogh}
%\int_{p_\alpha (t,z_{datum})}^{p_\alpha (t,z)} %\frac{dp_\alpha^{'}}{\rho_\alpha \left(p_\alpha^{'}\right)} = g\left(z - %z_{datum}\right) \, ,
%\end{equation}

%where $p_\alpha (t,z)$ is the pressure field of phase $\alpha$ and $p_\alpha (t,z_{datum})$ is the pressure at the datum level $z_{datum}$. If the pressure at the datum level is known, it is possible to compute the pressure field from Eq.~\eqref{eq: p=rhogh} and average fluid properties accordingly.

It will be shown that the pressure field can be further simplified by neglecting the variation of density with depth. For instance, the average oil density can be approximated to the local density at the datum level, $\rho_o (t,z) \approx \rho_o (t, z_{datum})$, what results in the following simple expression for the pressure field:
\begin{equation}
p_\alpha (t,z) = p_\alpha \left(t, z_{datum}\right)+ \rho_\alpha (t, z_{datum}) g \left(z-z_{datum}\right) \, .
\end{equation}

@@

 As pressure increases with depth according to the hydrostatic gradient, density also increases but by a negligible amount. This approximation can be quantified by putting together the expressions for compressibility and hydrostatic equilibrium. For this, we assume a constant isothermal compressibility for oil phase,
\begin{equation}\label{eq: grado}
\frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial p_o} \right)_{T,Y_o} = c_o \, ,
\end{equation}
 and we use hydrostatic equilibrium $dp_o = \rho_o g dz$ to obtain the expression for the derivative of oil density with depth:
\begin{equation}\label{eq: gradodz}
 \frac{1}{\rho_o}\left(\frac{\partial \rho_o}{\partial z} \right)_{T,Y_o} = c_o \rho_o g \, .
\end{equation}

@@

Assuming reservoir temperature to be sufficiently above gas phase critical temperature and thus a negligible variation of the Z-factor with pressure, we use hydrostatic equilibrium again and obtain the derivative of gas density with depth:
\begin{equation}\label{eq: gradgdz}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T} = \frac{\rho_g \vec{g}}{p_g} = \left(\frac{\mathcal{M}}{ Z R T}\right) g \, .
\end{equation}

@@

Oil and gas densities are thus approximately constant with depth and their average values can be approximated to the local values at the datum level. The pressure calculated in the material balance equation thus corresponds to the local pressure at $z_{datum}$.

@@

For the case of an oil reservoir with a gas cap, it is more accurate to define the datum level at the gas-oil contact. The point is that a higher gas density will be computed for a datum level bellow gas-oil contact because of the higher hydrostatic gradient of oil phase,
\begin{equation}
\frac{1}{\rho_g}\left(\frac{\partial \rho_g}{\partial z}\right)_{T,Z} = \frac{\rho_o}{\rho_g}\left(\frac{\mathcal{M}}{Z R T}\right)g  \, ,
\end{equation}
what is $\rho_o / \rho_g$ times greater than the gas hydrostatic gradient used in Eq.~\eqref{eq: gradgdz}.

It was shown that the pressure variation with depth can be neglected for the computation of gas, oil and water densities. This is true provided that phase changes do not occur.

 for what the pressure variation with depth may not be negligible in the solubility ratio: 
\begin{equation}\label{eq: dRsdz1}
\frac{1}{R_s}\frac{d R_s}{dz} = \frac{1}{R_s}\frac{d R_s}{dp_o} \frac{dp_o}{dz} = \frac{\rho_o g}{R_s}\frac{dR_s}{dp_o} \, .
\end{equation}

@@

\section{Solution of material balance equation}
A numerical solution of Eq.~\eqref{eq: MBE_short} is presented here. We start by defining a residual function ($\varepsilon$) for which the pressure solving Eq.~\eqref{eq: MBE_short} is a root:
\begin{equation}\label{eq: residual}
\varepsilon(p_2) = \frac{E+F_i+C-F_p}{V_{p1}} \, .
\end{equation}
We proceed showing that $\varepsilon$ is a monotonic function provided some conditions on fluid properties are observed and that, as a consequence, only one root of $\varepsilon$ exists at most. Equivalently, we can say that only one solution exists for the material balance equation.

@@

The residual function clearly satisfies the first two affirmatives because phase properties are continuous everywhere and are also differentiable everywhere except for the phase change at bubble-point pressure.
The third affirmative is proved by rewriting $d\epsilon/dp$ as a sum of four parcels, each corresponding to a different phase (three fluid phases and the solid rock). For the fluid phases, each parcel is composed of two factors: current volume \textit{in-situ} and volume derivative with respect to pressure:
\begin{equation}
\begin{split}
&V_{p1}\frac{d\varepsilon}{dp}=\underbrace{\left(N-N_p\right) \left(\frac{dB_o}{dp} - B_g \frac{dR_s}{dp} \right)}_\text{oil phase}\\
&+\underbrace{\left[N R_{s1}+mN\frac{B_{o1}}{B_{g1}}+G_i-\left(N-N_p\right)R_{s2}-G_p\right] \frac{dB_g}{dp}}_\text{gas phase}\\
&+\underbrace{\left(W + W_i+W_e-W_p\right)\frac{dB_w}{dp}}_\text{water phase}\\
&\underbrace{-\frac{dV_p}{dp}}_\text{porous volume}\, .
\end{split}
\end{equation}

\textit{In-situ} volumes (e.g. $N-N_p$ for oil phase) are non-negative values,  pressure derivatives of fluid volumes are non-positive values (fluid volume cannot increase as pressure increases) and the pressure derivative of porous volume is a positive value (as fluid pressure increases, porous volume expands owing to rock compression).

The value of $d\varepsilon/dp$ is thus negative if at least one phase is compressible because it is a sum of products between negative and positive factors.

@@

It must be observed that the result of gas reentering oil phase completely neglects non-equilibrium effects, e.g. compositional gradients and diffusion of matter, and must be used with care. .
